SunPy - Python for Solar Physics

### 2014-01-30 4-Util.tex
comments and some little fixings
Edits to the rest

\subsection{World Coordinate System (WCS) Coordinates}\label{ssec:util:wcs}
Coordinate transformations are frequently a necessary task within the solar 
data analysis workflow. Likely the most often used transformation is from 
observer coordinates (e.g., sky coordinates) to a coordinate system which is 
mapped onto the solar surface (e.g., latitude and longitude). This 
transformation is necessary to compare the true physical distance between 
different solar features. This type of transformation is not unique
to solar observations, but is not often considered by astronomical packages
such as the Astropy 
coordinates package. The \verb|wcs| package in SunPy implements the World Coordinate 
System (WCS) for solar coordinates as described by \cite{Thompson2000}. The 
transformations currently implemented are those most useful 
for most solar data analysis, namely converting from Helioprojective-Cartesian 
(HPC) to Heliographic (HG) coordinates. HPC describes the positions on 
the Sun as angles measured from the center of the solar disk (usually in 
arcseconds) using cartesian coordinates (X, Y). This is the coordinate system 
most often defined in solar imaging data (see for example, images from 
\textit{SDO/AIA}, \textit{SOHO/EIT}, and \textit{TRACE}). 
HG coordinates express positions on the Sun using longitude and latitude on 
the solar sphere. There are two standards for this coordinate system, 
Stonyhurst-Heliographic where the origin is at the intersection of the solar 
equator and the central meridian as seen from Earth and 
Carrington-Heliographic is a system which is fixed to the Sun and does not depend on Earth. The 
implementation of these transformations pass through a common coordinate system 
called Heliocentric-Cartesian (HCC) where positions are expressed in true 
(deprojected) physical distances instead of angles on the celestial sphere.
These transformations require some knowledge of the location of the observer 
which is usually provided by the image header. In the case where it is 
not, the observer is assumed to be at Earth. Listing \ref{code:wcs_code} shows 
some examples of coordinate transforms carried out in SunPy using the 
\verb|wcs| utilities.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import wcs
>>> wcs.convert_hg_hpc(10,53)
(100.49244115330731, 767.97438321917502)
# Convert that position back to heliographic coordinates
>>> wcs.convert_hpc_hg(100.49, 767.97)
(9.9996521808465175, 52.999563684874893)
# Try to convert a position which is not on the Sun to HG
>>> wcs.convert_hpc_hg(-1500,0)
sunpy/wcs/wcs.py:180: RuntimeWarning: invalid value encountered in sqrt
  distance = q - np.sqrt(distance)
(nan, nan)
Convert sky coordinate to a position in HCC
>>> wcs.convert_hpc_hcc(-300,400, z = True)
(-216716967.63331246, 288956420.9477042, 594364636.2208252)
\end{minted}
\caption{Using the sunpy.wcs subpackage.}
\label{code:wcs_code}
\end{listing}

\subsubsection{Sun}\label{ssec:util:sun}
The purpose of this module is to provide solar-specific data, for example ephemerides and
solar constants. The main namespace contains a number of functions which provide solar
ephemerides such as the Sun to Earth distance, solar cycle number, the mean 
anomaly, etc.
All of these functions take a time as their input which can be provided in any 
\texttt{parse\_time}
compatible format. 
A future version of SunPy will integrate with the Astropy coordinates package and will 
likely make use of an ephemeris package.

\subsubsection{constants}\label{ssec:util:const}
The purpose of the constants package is to provide a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a Constant object as defined by Astropy. Each 
Constant object defines a Quantity, a number associated with a unit, along with 
the constant's provenance (i.e., reference) as well as its uncertainty. Using 
Astropy's Quantities, any solar constant can easily be converted between 
different units as can be seen in Listing~\ref{code:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
Numpy's \texttt{ndarray}, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the sub-package. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly a \texttt{print\_all()} function is available.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.sun import constants as sun
>>> print(sun.mass)
  Name   = Solar mass
  Value  = 1.9891e+30
  Error  = 5e+25
  Units  = kg
  Reference = Allen's Astrophysical Quantities 4th Ed.
# Verify the average density of the Sun and convert to cgs
>>> (sun.mass/sun.volume).cgs
<Quantity 1.40851154227 g / (cm3)>
# Search for the age of the Sun
>>> sun.find('age')
['age', 'average angular size', 'average density', 'average intensity']
>>> sun.value('age'), sun.unit('age')
(4600000000.0, Unit("yr"))
\end{minted}
\caption{Using the sunpy.sun.constants sub-package.}
\label{code:constants_code}
\end{listing}

\subsubsection{Instruments}\label{ssec:util:inst}
In addition to providing support for instrument-specific solar data via the main data 
classes \texttt{Map}, \texttt{LightCurve} and \texttt{Spectrum}, 
some instrument-specific functions may be found within the \texttt{instr} package. 
These functions are generally those that are unique to one particular solar instrument, 
rather than of general use, for example a function to construct a GOES flare event list 
or a function to query the LYRA timeline annotation file. Currently, some support is included
 for the GOES, LYRA, RHESSI and IRIS instruments, while future developments 
 will include support for additional missions. Ultimately, it is anticipated that solar
  missions requiring a large suite of software tools will be supported via a separately 
  maintained package affiliated with SunPy.

>>>>

\subsection{World Coordinate System (WCS) Coordinates}\label{ssec:util:wcs}
Coordinate transformations are frequently a necessary task within the solar 
data analysis workflow. Likely the most often used transformation is from 
observer coordinates (e.g., sky coordinates) to a coordinate system which is 
mapped onto the solar surface (e.g., latitude and longitude). This 
transformation is necessary to compare the true physical distance between 
different solar features. This type of transformation is not unique
to solar observations, but is not often considered by astronomical packages
such as the Astropy 
coordinates package. The \texttt{wcs} package in SunPy implements the World Coordinate 
System (WCS) for solar coordinates as described by \citep{thompson2006}. The 
transformations currently implemented are those most useful 
for most solar data analysis, namely converting from Helioprojective-Cartesian 
(HPC) to Heliographic (HG) coordinates. HPC describes the positions on 
the Sun as angles measured from the center of the solar disk (usually in 
arcseconds) using Cartesian coordinates (X, Y). This is the coordinate system 
most often defined in solar imaging data (see for example, images from 
\textit{SDO}/AIA, \textit{SOHO}/EIT, and \textit{TRACE}). 
HG coordinates express positions on the Sun using longitude and latitude on 
the solar sphere. There are two standards for this coordinate system:
Stonyhurst-Heliographic, where the origin is at the intersection of the solar 
equator and the central meridian as seen from Earth, and 
Carrington-Heliographic, which is fixed to the Sun and does not depend on Earth. The 
implementation of these transformations pass through a common coordinate system 
called Heliocentric-Cartesian (HCC), where positions are expressed in true 
(de-projected) physical distances instead of angles on the celestial sphere.
These transformations require some knowledge of the location of the observer 
which is usually provided by the image header. In the case where it is 
not, the observer is assumed to be at Earth. Listing \ref{code:wcs_code} shows 
some examples of coordinate transforms carried out in SunPy using the 
\texttt{wcs} utilities.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import wcs
>>> wcs.convert_hg_hpc(10,53)
(100.49244115330731, 767.97438321917502)
# Convert that position back to heliographic coordinates
>>> wcs.convert_hpc_hg(100.49, 767.97)
(9.9996521808465175, 52.999563684874893)
# Try to convert a position which is not on the Sun to HG
>>> wcs.convert_hpc_hg(-1500,0)
sunpy/wcs/wcs.py:180: RuntimeWarning: invalid value encountered in sqrt
  distance = q - np.sqrt(distance)
(nan, nan)
Convert sky coordinate to a position in HCC
>>> wcs.convert_hpc_hcc(-300,400, z = True)
(-216716967.63331246, 288956420.9477042, 594364636.2208252)
\end{minted}
\caption{Using the \texttt{wcs} subpackage.}
\label{code:wcs_code}
\end{listing}

\subsubsection{Sun}\label{ssec:util:sun}
The purpose of the \texttt{sun} subpackage is to provide solar-specific data such as ephemerides and
solar constants. The main namespace contains a number of functions which provide solar
ephemerides such as the Sun to Earth distance, solar-cycle number, the mean 
anomaly, etc.
All of these functions take a time as their input which can be provided in a format
compatible with \texttt{sunpy.time.parse\_time()}. 
A future version of SunPy will integrate with the Astropy coordinates package and will 
likely make use of an ephemeris package.

%ays: I removed this as a separate subsection because constants is part of sun...
%\subsubsection{constants}\label{ssec:util:const}
The \texttt{sun.constants} module provides a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a \texttt{Constant} object as defined by Astropy. Each 
\texttt{Constant} object defines a \texttt{Quantity}, a number associated with a unit, along with 
the constant's provenance (i.e., reference) as well as its uncertainty. Using 
Astropy's {Quantity} objects, any solar constant can easily be converted between 
different units as can be seen in Listing~\ref{code:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
NumPy's \texttt{ndarray}, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the module. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly, a \texttt{print\_all()} function is available.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.sun import constants
>>> print(sun.mass)
  Name   = Solar mass
  Value  = 1.9891e+30
  Error  = 5e+25
  Units  = kg
  Reference = Allen's Astrophysical Quantities 4th Ed.
# Verify the average density of the Sun and convert to cgs
>>> (constants.mass/constants.volume).cgs
<Quantity 1.40851154227 g / (cm3)>
# Search for the age of the Sun
>>> constants.find('age')
['age', 'average angular size', 'average density', 'average intensity']
>>> constants.value('age'), constants.unit('age')
(4600000000.0, Unit("yr"))
\end{minted}
\caption{Using the \texttt{sun.constants} module.}
\label{code:constants_code}
\end{listing}

\subsubsection{Instruments}\label{ssec:util:inst}
In addition to providing support for instrument-specific solar data via the main data 
classes \texttt{Map}, \texttt{LightCurve} and \texttt{Spectrum}, 
some instrument-specific functions may be found within the \texttt{instr} subpackage. 
These functions are generally those that are unique to one particular solar instrument, 
rather than of general use, such as a function to construct a \textit{GOES} flare event list 
or a function to query the \textit{LYRA} timeline annotation file. Currently, some support is included
 for the \textit{GOES}, \tetxit{LYRA}, \textit{RHESSI} and \textit{IRIS} instruments, while future developments 
 will include support for additional missions. Ultimately, it is anticipated that solar
  missions requiring a large suite of software tools will be supported via a separately 
  maintained package that is affiliated with SunPy.

### 2014-01-30 solarsoft.tex
Edits to the rest

% RJH: The entire next two paragraphs never mention SSW.  I find all of the
% lines of code and $$ counts to be superfluous.  Do those include the documentation?
% What about generated files?  I don't think they contribute much.

SSW relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis environment 
sold by Exelis, a global aerospace, defence, information and services company. 
The SSW `gen' package provides the base analysis capabilities such as time-series
analysis, time conversions, spectral fitting, image display, and file 
I/O. It is composed of 89\% IDL code while the remainder is mostly Perl (5.4\%) 
and csh scripts (3.8\%). 
According to data generated using David A. Wheeler's `SLOCCount'\footnote{\url{http://www.dwheeler.com/sloccount/}}, SSW/gen is composed of 
289,724 lines of code with a total approximate value of \$10.4M (based on the 
Basic COCOMO model \citep{cocomo}). 
%dps - This data was calculated using...} If we can add them to the repository 
%will improve the quality of the paper and we can always repeat the calculation in the future.
%schriste - i don't know what you mean by this 
%Stuart: comment was Davids
Individual missions and instruments provide additional optional packages to 
support and process their data. For example, the \texttt{sdo} package \citep{sdo} is 
100\% IDL code and is 11k lines of code (\$360k) and the \texttt{soho} package 
\citep{soho} is 100k lines of code (\$3M) and is 99.9\% IDL code. SSW is hosted 
and distributed by the Lockheed Martin Solar and Astrophysics Laboratory.
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW itself is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.
Even though SSW's source is open, the development of the source is primarily closed,
with no open issue tracker or established mechanism for community contributions.

One of SunPy's key aims is to provide a free and modern alternative to the 
SolarSoft (gen) library. This is made possible by the recent rise of the 
scientific Python community. Python, as described in Section \ref{sec:Intro}, 
is a general purpose high-level programming language which is free to use and 
cross-platform and powerful programming language.
It therefore has a very large user base which extends far outside of the 
solar physics or astronomy communities. It is very extensible with C, C++, 
FORTRAN or even IDL. Many books and on-line documentation sources are available. For 
example, at the time writing a search 
%
%dps - \footnote{We need to be careful with this! IDL means more things that 
%our 
%beloved Interactive Data Language. If we search for that and programming 
%then we get just 7 books! - from which 5 are just the same. I don't know how we 
%could make a more accurate search.
%If we script it it would be amazing!  Also, I would just provide the order
%of magnitude, and not a exact number, i.e., 190 and 1800 results}
% 
% schriste - I think we defined IDL properly. I was being generous with the 
%book search for IDL
%Stuart: again not not mine, I think we should drop the book thing, in 
%comparison to the other metrics used it's a little weak.
on Amazon for books on IDL programming generated $189$ results while a search 
for Python yielded $1\,788$ results. Python has a strong emphasis on 
readability, a concept which is well-aligned with the scientific endeavour.
%In fairness, IDL is also quite readable! Could move this statement to the intro instead.
Additionally, Python is increasingly being taught in universities, which means 
that research projects can begin quickly without the need for students to learn 
a new (and relatively niche) language. Finally, knowledge of Python is a skill 
which is useful and well compensated outside of the scientific community 
(average salary of \$110k in the Washington, DC, area according to indeed.com at 
% RJH: Is this for a python developer?  Again, this is useless info, imho.
the time of writing). Python continues to see increased use in the astronomy 
community \citep{greenfield2011} which has similar goals and requirements as 
the solar physics community. Finally, Python integrates well with many 
technologies such as web servers \citep{dolgert2008}, SQL databases, 
interactive and shareable notebook-based computing \citep{perez2007}. For all 
of these reasons, Python is an ideal platform for a modern data analysis environment 
for solar physics.

SunPy is built on top of the core scientific Python packages namely 
\texttt{NumPy}, \texttt{SciPy}, \texttt{matplotlib}. These core packages 
provide capabilities surpassing that provided by IDL for array manipulation and 
2D plotting \citep{greenfield2011}. Providing similar capabilities to SSW/gen 
in Python is made easier by leveraging the diverse Python community. For 
example, SunPy makes use of the \texttt{pandas} package to provide 
high-performance, easy-to-use data structures and data analysis tools for time 
series data \citep{mckinney2012}. \texttt{pandas} was originally developed for 
quantitative analysis of financial data and has since grown into a generalised 
data analysis package. SunPy also relies on and is coordinating with the 
\texttt{astropy} package \citep{theastropycollaboration2013}. Astropy is still in 
active development but already provides many relevant capabilities such as 
full-featured support for FITS files, absolute date and time conversions, 
units and unit conversions, and model-fitting. Pandas (134k lines of code) 
and Astropy (90k lines of code) allow SunPy to leverage a total value of \$7.6M 
worth of development which is close to the \$10.4M for SSW/gen assuming a rough 
equivalency between a line of Python code and a line of IDL code.

>>>>

% RJH: The entire next two paragraphs never mention SSW.  I find all of the
% lines of code and $$ counts to be superfluous.  Do those include the documentation?
% What about generated files?  I don't think they contribute much.
% dps: I agree... I don't like the lines of code connection with work or efficiency
%   but at the same time is something used by some people... so... I don't know.

SSW relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis environment 
sold by Exelis, a global aerospace, defence, information and services company. 
The SSW \texttt{gen} package provides the base analysis capabilities such as time-series
analysis, time conversions, spectral fitting, image display, and file 
I/O. It is composed of 89\% IDL code while the remainder is mostly Perl (5.4\%) 
and csh scripts (3.8\%). 
According to data generated using David A. Wheeler's `SLOCCount'\footnote{\url{http://www.dwheeler.com/sloccount/}}, SSW/\texttt{gen} is composed of 
289,724 lines of code with a total approximate value of \$10.4M (based on the 
Basic COCOMO model \citep{cocomo}). 
%dps - This data was calculated using...} If we can add them to the repository 
%will improve the quality of the paper and we can always repeat the calculation in the future.
%schriste - i don't know what you mean by this 
%dps: schriste, could you tell -at least us- how that was measured? can we make an
% script that we could use next year and see how much it changed?
Individual missions and instruments provide additional optional packages to 
support and process their data. For example, the \texttt{sdo} package \citep{sdo} is 
100\% IDL code and is 11k lines of code (\$360k) and the \texttt{soho} package 
\citep{soho} is 100k lines of code (\$3M) and is 99.9\% IDL code. SSW is hosted 
and distributed by the Lockheed Martin Solar and Astrophysics Laboratory.
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW itself is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.
Even though SSW's source is open, the development of the source is primarily closed,
with no open issue tracker or established mechanism for community contributions.

One of SunPy's key aims is to provide a free and modern alternative to the 
SSW/\texttt{gen} library. This is made possible by the recent rise of the 
scientific Python community. Python, as described in Section \ref{sec:Intro}, 
is a general purpose high-level programming language which is free to use and 
cross-platform and powerful programming language.
It therefore has a very large user base which extends far outside of the 
solar physics or astronomy communities. It is very extensible with C, C++, 
FORTRAN or even IDL. Many books and online documentation sources are available. For 
example, a simple search on Amazon.com yields hundreds of books on Python programming.
%
%dps - \footnote{We need to be careful with this! IDL means more things that 
%our 
%beloved Interactive Data Language. If we search for that and programming 
%then we get just 7 books! - from which 5 are just the same. I don't know how we 
%could make a more accurate search.
%If we script it it would be amazing!  Also, I would just provide the order
%of magnitude, and not a exact number, i.e., 190 and 1800 results}
% 
% schriste - I think we defined IDL properly. I was being generous with the 
%book search for IDL
%Stuart: again not not mine, I think we should drop the book thing, in 
%comparison to the other metrics used it's a little weak.
% schriste - ok, how about this?
%dps: sounds better... but I'm skeptic of using Amazon.com...
Python has a strong emphasis on readability, a concept which is well-aligned with scientific endeavours.
%In fairness, IDL is also quite readable! Could move this statement to the intro instead.
Additionally, Python is increasingly being taught in universities, which means 
that research projects can begin quickly without the need for students to learn 
a new (and relatively niche) language. Finally, knowledge of Python is a skill 
which is useful and well compensated outside of the scientific community 
(average salary of \$110k in the Washington, DC, area according to indeed.com at 
% RJH: Is this for a python developer?  Again, this is useless info, imho.
% schriste - it is the only way to make a comparison and is fairly accepted by
% the software engineering community as far as i can tell.
the time of writing). Python continues to see increased use in the astronomy 
community \citep{greenfield2011} which has similar goals and requirements as 
the solar physics community. Finally, Python integrates well with many 
technologies such as web servers \citep{dolgert2008}, SQL databases, and 
interactive and shareable notebook-based computing \citep{perez2007}. For all 
of these reasons, Python is an ideal platform for a modern data analysis environment 
for solar physics.

SunPy is built on top of the core scientific Python packages namely 
\texttt{NumPy}, \texttt{SciPy}, \texttt{matplotlib}. These core packages 
provide capabilities surpassing that provided by IDL for array manipulation and 
2D plotting \citep{greenfield2011}. Providing similar capabilities to SSW/\texttt{gen} 
in Python is made easier by leveraging the diverse Python community. For 
example, SunPy makes use of the \texttt{pandas} package to provide 
high-performance, easy-to-use data structures and data analysis tools for time 
series data \citep{mckinney2012}. \texttt{pandas} was originally developed for 
quantitative analysis of financial data and has since grown into a generalised 
data analysis package. SunPy also relies on and is coordinating with the 
\texttt{astropy} package \citep{theastropycollaboration2013}. \texttt{astropy} is still in 
active development but already provides many relevant capabilities such as 
full-featured support for FITS files, absolute date and time conversions, 
units and unit conversions, and model-fitting. \texttt{pandas} (134k lines of code) 
and \texttt{astropy} (90k lines of code) allow SunPy to leverage a total value of \$7.6M 
worth of development, which is close to the \$10.4M for SSW/\texttt{gen} assuming a rough 
equivalency between a line of Python code and a line of IDL code.
